Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-10-25 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Updated the introduction

A world without any variety is barely conceivable. It certainly would not admit scientific investigation. We recall, though we have been unable to relocate, a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable
at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.

Without committing ourselves to Leibniz's overarching principle, our aim in this paper is to draw attention to the specific nature of the variety that can be observed in the universe and the unique suitability of Shannon's entropy (or information) formula $-\sum_ip_i\,\textrm{ln}\,p_i$ for the characterization of the variety. Indeed, as we will show in the next section, opportunities for application of Shannon's formula are ubiquitous both on the earth and in the universe at large.

We also wish to comment on the various interpretations and uses of Shannon's formula and attempt to answer this question: what does the formula represent? It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Botlzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. Indeed, it is often claimed that Shannon's formula is epistemic in nature and represents ignorance. There certainly are situations in which that is true but equally there are others in which the use of the formula is very appropriate and no ignorance is involved at all. Examples will be given in the next section.

This leads us to propose a different characterization of Shannon's formula that is both more general and more appropriate. Indeed, we will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements} and we will use the expression Shannon ``variability'', rather than ``entropy'' or ``information''. Moreover, the formula has a very wide range of applicability because the universe contains many different kinds of things that, through having common attributes, can be recognized as belonging together despite the fact that their attributes are expressed to different degrees or in different ways. To give a first example, the girls in a school can differ in the length and colour of their hair. More examples will follow in the next section. In this paper we do not ask why the universe sustains the variety that it does. We take it as a given.

We conclude this section with a word on terminology. In mathematics, Cantor defined a set as ``a collection into a whole $M$ of definite and separate objects''. He called the objects \emph{elements}. We will also use the word `elements' to denote the things we consider but subject to the proviso that they have a distinguishing feature (or features) that allows us to say that \emph{they are of the same kind}. Thus, we consider sets of a restricted type. The point is this. To constitute a set in accordance with Cantor's definition, it is only necessary for its elements to be \emph{definite and separate} objects. This leaves total freedom as to how such objects in the universe are to be parcelled up into specific sets. For example, a set might consist of Mars and the word `go'. We think such latitude misses one of the most significant aspects of the universe, namely, that the things within it can be distinguished by attributes that are much more specific than the two required by Cantor. Our field of investigation goes beyond mere number, which washes out all attributes apart from the two, of being definite and separate, that allow us to count. They were the only attributes that Cantor needed; we need more.

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. What we would like to say is that the variability of the overall distribution $H_T = H_C + p_M H_M$ will be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity linear when combining two independent distributions.

>>>>

The use of Shannon entropy $-\sum p_i \log p_i$ is well established in very different fields of knowledge such as computer science[cite], physics[cite] and ecology[cite]. It is a fundamental element in information theory since Shannon developed his mathematical theory of communication. It has been used extensively within statistical mechanics to link the microscopic description of a system to its macroscopic thermodynamic one. It has been used in ecology to characterize biodiversity and calculate species distributions under given constraints. Entropy maximization has been shown to be a powerful tool in all these settings.

While the techniques are well developed, it is puzzling that this concept can be applied in such different contexts, given that it assumes different features in each one. For example, in information theory the Shannon entropy for a book will change whether we choose to encode words or letters, which means the quantity does not have a fully objective feature: it depends on an arbitrary choice. This is not problematic for an information theoretic quantity, but it is for a physical quantity. In statistical mechanics the $p_i$ represent probabilities, which means the Shannon entropy is a property of an ensemble of physical systems. This leads some to claim that the Shannon entropy is epistemic in nature and represents ignorance. [Add references] In ecology, however, the $p_i$ represent the fraction of individuals belonging to the $i$th species. There is no probabilistic ensemble to which one can associate a degree of ignorance. Finally, thermodynamic entropy is associated to the notion of states and irreversible time evolution, which are not well defined concepts for the other fields. The same expression, then, seems to have acquire different flavors and it is not clear why a single concept would apply in all these cases.

To make matter worse, it is fairly typical to describe entropy with notions such as uncertainty[cite], knowledge[cite], lack of knowledge[cite], disorder[cite]. While some of these characterizations are useful in some contexts to give an intuitive picture, they do not seem to always work. A common objection to the characterization as disorder, for example, is that a uniform distribution corresponds to a homogeneous system, which one would expect to be maximally ordered; yet the distribution maximizes the Shannon entropy, which would indicate maximal disorder. Moreover, these notions remain vague in the sense that the formula is not formally derived by those concepts and therefore it is not clear why that particular indicator would be an appropriate choice. We are left to wonder: is there really some single underlying concept, that, when applied appropriately in the different fields, reduces to the correct notion of entropy? Or is it just a big coincidence?

We propose that there is indeed a characterization of Shannon's formula that is both more general and more appropriate; that is independent on whether it is applied to sources of messages, ensembles of microstates or animal populations. We will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements}. That is, given a distribution of some elements over some quantity, the Shannon entropy represents how varied, how different from each other, the values are. In fact, it is the only continuous and linear index that measures the degree of variability within a distribution, and those requirements lead naturally to same three conditions that Shannon set for his formula. In other words, the concept of variability with a distribution leads in a formal way to the correct expression. We will then show how this perspective naturally connects to information theory and statistical mechanics when applied in the respective contexts.

As we believe this is the right characterization, we will therefore use the expression Shannon ``variability'', rather than ``entropy'' or ``information'', as it tells us exactly what the quantity describes. It is well known that when Shannon recognized the importance of his formula in communication theory he sought von Neumann's advice on the name he should give to the expression he had found. He was told to call it \emph{entropy}, first, because Boltzmann had used an identical formula to define entropy in statistical mechanics and, second, nobody understood entropy and that would provide protection from criticism [exact words to be found]. We believe von Neumann's advice was not totally helpful and has given rise to confusion. We will reserve the word entropy for the thermodynamic entropy, the one dealing with physical states and irreversible processes.

This characterization also explains why the Shannon variability has such a broad range of applicability. As nature presents itself in a multitude of forms and combinations, a fundamental task in science is to recognize which objects belong together and study how their different attributes are expressed in different ways or degrees. They may galaxies, animals or outcomes of the same process; we may study their shape, behavior or numerical quantities. But the idea of grouping them and try to understand the source of their variation is a common.\footnote{In fact, determining which collection of objects is significant and what attributes are relevant is significant part of the work.} Measuring and explaining this variability is an essential task. It is this very variety that is the inspiration for and the subject of scientific investigation. A universe without variety, where all objects were the same and remained the same, is barely conceivable and certainly it would leave no space for scientific investigation.

As a final note, we want to stress that the goal of this work is not to present some new technical result. The goal is to put all details together in a coherent picture that gives them context and see precisely how they are related. This can help, especially the new practitioners, to avoid common misconceptions. Moreover, while some details presented here may be well known in a particular community, we find that it may be new to another. For example, we will show an example of Huffmann coding to illustrate the meaning of the numerical value associated with the Shannon variability. This is standard practice in computer science courses, yet it is not in statistical mechanics. Therefore restating all the crucial aspects in one place is beneficial because the connection between them is not always readily available.

%https://www.researchgate.net/publication/259011656_A_tribute_to_Claude_Shannon_1916-2001_and_a_plea_for_more_rigorous_use_of_species_richness_species_diversity_and_the_'Shannon-Wiener'_Index

%https://web.stanford.edu/~hastie/Papers/maxent_explained.pdf

%https://www.mdpi.com/1099-4300/21/8/794/htm

%Leibniz
%We recall, though we have been unable to relocate, a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''. [REF]

%We conclude this section with a word on terminology. In mathematics, Cantor defined a set as ``a collection into a whole $M$ of definite and separate objects''. He called the objects \emph{elements}. We will also use the word `elements' to denote the things we consider but subject to the proviso that they have a distinguishing feature (or features) that allows us to say that \emph{they are of the same kind}. Thus, we consider sets of a restricted type. The point is this. To constitute a set in accordance with Cantor's definition, it is only necessary for its elements to be \emph{definite and separate} objects. This leaves total freedom as to how such objects in the universe are to be parcelled up into specific sets. For example, a set might consist of Mars and the word `go'. We think such latitude misses one of the most significant aspects of the universe, namely, that the things within it can be distinguished by attributes that are much more specific than the two required by Cantor. Our field of investigation goes beyond mere number, which washes out all attributes apart from the two, of being definite and separate, that allow us to count. They were the only attributes that Cantor needed; we need more.

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the probability. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ will be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity linear when combining two independent distributions.
